---
title: "Reproducible Research: Peer Assessment 1"
---

```{r setoptions, echo=FALSE, warning = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE )
```
## Loading and preprocessing the data
```{r}
if(!file.exists("activity.csv")) {
         file <- unzip("activity.zip")
}

data <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
```{r}
stepsPerDay <- tapply(data$steps, data$date, sum, na.rm = TRUE)

```
  
  
#### Histogram of the total number of steps taken each day  

```{r}
library(ggplot2)
qplot(stepsPerDay, xlab = "Total Steps", ylab = "Frequency", binwidth = 500)
```
  

#### Mean and median number of steps taken each day  

```{r}
meanStepsPerDay <- mean(stepsPerDay)
medianStepsPerDay <- median(stepsPerDay)
```
  
* Mean: 9354.2295
* Median:  10395

-----  

## What is the average daily activity pattern?  

```{r}
average <- aggregate(x=list(steps=data$steps),by=list(interval=data$interval), FUN = mean, na.rm = TRUE)

```
  
#### Time series plot of the average number of steps taken  

```{r}

ggplot(data = average, aes(x=interval, y=steps)) +
geom_line() +
xlab("Interval") +
ylab("Average number of steps taken")
```
  
#### The 5-minute interval that, on average, contains the maximum number of steps  

```{r}
average[which.max(average$steps),]
```
  
* Most Steps at: 8:35

----  

## Imputing missing values  

Many days/intervals contain missing values. To impute missing values we first calculate the total number of missing values. 

```{r}
numMissingValues <- length(which(is.na(data$steps)))
```
  
* Number of missing values: 2304  

We then create a new dataset and fill in the missing values  

```{r}

activityDataImputed <- data
activityDataImputed$steps <- sapply(data$steps, mean)

stepsByDayImputed <- tapply(activityDataImputed$steps, activityDataImputed$date, sum)
```
  
#### Histogram of the total number of steps taken each day after missing values are imputed  

```{r}
qplot(stepsByDayImputed, xlab='Total steps per day', ylab='Frequency', binwidth=500)
```
  
#### The mean and median total number of steps taken per day  

```{r}
stepsByDayMeanImputed <- mean(stepsByDayImputed, na.rm = TRUE)
stepsByDayMedianImputed <- median(stepsByDayImputed)
```
  
* Mean (Imputed): 1.0766 &times; 10<sup>4</sup>
* Median (Imputed):  1.0766 &times; 10<sup>4</sup>


----  

## Are there differences in activity patterns between weekdays and weekends?  

At first we create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.  

```{r}
day <- function(date) {
  day <- weekdays(date)
  if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
    return("weekday")
  else if (day %in% c("Saturday", "Sunday"))
    return("weekend")
  else
    stop("invalid date")
}
data$date <- as.Date(data$date)
data$day <- sapply(data$date, FUN=day)

```
  
#### Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends   

```{r}
averages <- aggregate(steps ~ interval + day, data=data, mean)
ggplot(averages, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) +xlab("5-minute interval") + ylab("Average number of steps")
```